Non-word repetition in Yélî Dnye

### 2020-12-18 manuscript.Rmd
minor typo fixes

NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively apparent phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant]. Further, the languages it has been used with have historically had small to moderate phonological inventories [@maddieson2005correlating].
<!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past? Or is there a specific hypothesis you had in mind about how inventory size affects phonological development and, hence, NWR performance? I could come up with something, but I think you had an idea already and I don't want to step on it!-->

In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation. Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations. Third, these data give a first impression of how this task is responded to in a rarely-studied language and culture; we add to this by further investigating whether there are structured sources that account for individual variation in our sample.
<!--M2A: This is also a note to myself; it seems we could use some more references in the above!-->

@@

Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing, @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/). However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CVCV structure instead (e.g., 'ndapî' /'ṇṭæ.pɯ/). This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/). An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.'læ/), but these are limited to /æ/-. Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).
<!--
The following was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular. There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that, even when there is deisel available, are primarily reserved for medical transport. There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG. There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time. As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, is typically unavailable. Our data collection is therefore limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households situated close to schools to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biilik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

Previous work typically steers clear of articulartorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology. We ultimately used a relatively large number of items that would enable us to explor both variation in structural complexity and in more vs. less challenging sounds. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below). Naturally, designing the task in this way may render the study of individual variation within the population more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit. That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.

@@

- Does the cross-linguistic frequency of sounds predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation attributable to child age, sex, birth order, monolingual status, and/or parental education?

>>>>

NWR has been used to seek answers to a variety of theoretical questions, including what the links are between phonology, working memory, and the lexicon [@bowey2001nonword], and how extensively phonological constraints found in the lexicon affect online production [@gallagher2014identity]. NWR is also frequently used in applied contexts, notably as a diagnostic tool for language delays and disorders [@estes2007differences]. Since non-words can be generated in any language, it has attracted the attention of researchers working in multilingual and linguistically diverse environments, particularly in Europe [@meir2016disentangling;@is08042009language]. However, it has seldom been used outside of Europe and North America [with exceptions including @gallagher2014identity and @cristia2020infant]. Further, the languages it has been used with have historically had small to moderate phonological inventories [@maddieson2005correlating].
<!--M2A: Is the motivating idea here that application to YD is of interest because it's very different from what we've look at with NWR in the past? Or is there a specific hypothesis you had in mind about how inventory size affects phonological development and, hence, NWR performance? I could come up with something, but I think you had an idea already and I don't want to step on it!-->

In the present study, we use NWR to investigate the phonological development of children learning Yélî Dnye, an isolate language spoken in Papua New Guinea (PNG) that has a large and unusually dense phonological inventory. The study was designed to contribute to multiple aspects of our understanding of phonological development: First, we included a subset of non-word items with typologically rare and/or challenging sounds to ask whether these rare sounds are disadvantaged in perception and/or production, both in terms of overall repetition scores and patterns of mispronunciation. Second, we varied the length (in syllables) of non-words to investigate the impact of word length on repetition and mispronunciation; previous NWR research has shown variable effects of length between populations. Third, these data give a first impression of how this task is responded to in a rarely-studied language and cultural setting; we add to this contribution by further investigating whether there are structured sources that account for individual variation in our sample.
<!--M2A: This is also a note to myself; it seems we could use some more references in the above!-->

@@

Most words in Yélî Dnye are bisyllabic (~50%), with monosyllabic words (~40%) appearing most commonly after that, and with tri-and-above syllabic words appearing least frequently [~10%; based on >5800 lexemes in the most recent Yélî Dnye dictionary at time of writing; @levinsonYDgrammar]. The vast majority of syllables use a CV format. A small portion of the lexicon features words with a final closed syllable, limited to codas of -/m/, -/p/, or -/j/ (e.g., 'ndap' (Spondylus shell) /ṇṭæp/). However, in spontaneous speech an epenthetic /ɯ/ is often inserted after word-final coda, yielding a CV.CV structure instead (e.g., 'ndapî' /ṇṭæ.pɯ/). This process is used frequently with English loan words that have a coda (e.g., 'ponî' (phone) /'pɔ.nɯ/). An even smaller portion of the lexicon features words starting with a vowel (e.g., 'ala' (here) /æ.læ/), but these are limited to /æ/-. Finally, the lexicon features a handful of single-vowel grammatical morphemes (see @levinsonYDgrammar for details).

<!--
NOTE: The following paragraph was removed b/c it's less relevant, but it has been double-checked by MC Nov 2020 in case we decide to include it later:

@@

Laying 250 nautical miles off the coast of mainland PNG and surrounded by a barrier reef, transport to and from the island is both infrequent and irregular. There is no electricity system (we use solar panels) or motor vehicles on the island, save dinghys with outboard motors that are primarily reserved for medical transport when there is diesel available. There is a medical radio connection at the health clinic, located at the Catholic Mission, via which news and messages are often passed to and from mainland PNG. There is also a cellular tower on the island, though it is often inoperative, usually for multiple months or years at a time. As such, access to news and outside connections, including international phone calls and digital exchanges that require significant data transfer, is typically unavailable. Our data collection is therefore limited to the duration of the researchers' on-island visits.

Despite restricted outside contact, formal education is a priority for many Yélî families. A recent study surveying more than 40 parents of young children in this region of the island found that nearly all parents completed 6 or more years of education, with 65% of mothers and 49% of fathers completing the full range of locally available education (~9 years), with around half of those parents having pursued an additional 2+ years of education on other islands in the region or on mainland PNG [@casillas2020early]. While there are a handful of local schools around the island, these may be well out of walking distance for many children (i.e., more than 1 hour on foot or by canoe each day). It is very common for households that are situated close to a school to billet their school-aged relatives during the weekdays (sometimes the weekends too, if the child's home is very far) for long segements of each school year. Combined with practices of collective child guardianship within each close-knit hamlet, the practical consequence of this billeting for researchers is that adult consent can often come from a combination of aunts, uncles, adult cousins, and grandparents standing in for the child's biological parents. On top of this, child assent is culturally pertinent, as child independence is encouraged and respected from toddlerhood onward [@brownIPchildrearing].^[For these reasons, we were unable to record precise details for a few children, including child date of birth, parent age in years, and parent education.] Finally, as mentioned, the language of schooling in this region is English, and so the majority of teens and adults on the island speak some English---many speak it very fluently.

@@

In a basic NWR task, the participant listens to a production of a word-like form, such as /bilik/, and then repeats back what they heard without changing any phonological feature that is contrastive in the language. For instance, in English, a response of [bilig] or [pilik] would be scored as incorrect; a response [biːlik], where the vowel is lengthened without change of quality would be scored as correct, because English does not have a general feature of contrastive vowel length. There is some variation in how past NWR studies have designed the presentation procedure and structure of items. For example, while items are often presented orally by the experimenter [@torrington2015non], an increasing number of studies have turned instead to playing back pre-recorded stimuli in order to increase control in stimulus presentation [@brandeker2015language]. Additionally, while some studies have used 10-15 non-words, others have employed up to 46 unique items [@piazzalunga2019articulatory]. Authors also often modulate structural complexity, typically measured in terms of item length (measured in number of syllables) and/or syllable structure [open as opposed to closed syllables, @gallon2007non].

Previous work typically steers clear of articulartorily and/or acoustically challenging sounds, but we included some in our experiment to more adequately represent Yélî Dnye's phonology. We ultimately used a relatively large number of items that would enable us to explore both variation in structural complexity and in more vs. less challenging sounds. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children (see below). Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items. However, a review of previous work on individual variation suggested to us that effects of individual differences in NWR are relatively small, and would not be detectable with the sample size that we could collect during our short visit. That said, we contribute to the literature by also reporting descriptive analyses of individual variation that could potentially be integrated in meta-analytic efforts.

@@

- Does the cross-linguistic frequency of sounds in the stimuli predict NWR scores? Are rarer sounds more often substituted by commoner sounds?
- How does score change as a function of item length in number of syllables?
- Is individual variation in NWR scores attributable to child age, sex, birth order, monolingual status, and/or parental education?
